---
title: Get started with bosc
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Get started with bosc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-lapis');});</script>

---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
if (requireNamespace("ggplot2", quietly = TRUE)) ggplot2::theme_set(ggplot2::theme_minimal())
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

## What does bosc do?

You have behavioral time series — reaction times, button presses, spike
trains — and you want to know whether they contain rhythmic structure. Is
there a dominant oscillation frequency? Is it statistically significant? At
what phase of the oscillation do events occur?

**bosc** answers these questions with a small, focused API:

| Verb | Function | What it does |
|:-----|:---------|:-------------|
| **Score** | `oscillation_score_z()` | Quantify oscillation strength + significance |
| **Aggregate** | `aggregate_by_bin()` | Turn trial data into regular subject-by-bin series |
| **Group spectrum** | `group_spectrum_test()` | Test rhythmic power across subjects |
| **Group TFR** | `group_tfr_test()` | Map where rhythmic power is concentrated |
| **Group phase** | `group_phase_consistency()` | Test phase alignment across subjects |
| **Filter** | `narrowband_hilbert()` | Extract phase and amplitude at a frequency |
| **Phase** | `phase_at_events()` | Get instantaneous phase at event times |
| **Cluster** | `detect_clusters()` | Find significant regions in time-frequency maps |
| **Test** | `circ_rayleigh()` | Test for non-uniform phase distributions |

```{r load-bosc}
library(bosc)
```

```{r grouped-demo-data, include=FALSE}
set.seed(7)
fs_group <- 30
n_subjects <- 8
n_bins <- 90
n_rep <- 4
nominal_bins <- seq(200, by = 1000 / fs_group, length.out = n_bins)
t_group <- seq(0, by = 1 / fs_group, length.out = n_bins)
envelope <- exp(-0.5 * ((seq_len(n_bins) - 50) / 11)^2)

trial_list <- vector("list", n_subjects * n_rep)
row_i <- 0L
for (subject_i in seq_len(n_subjects)) {
  for (rep_i in seq_len(n_rep)) {
    row_i <- row_i + 1L
    true_signal <- 0.9 * sin(2 * pi * 6 * t_group) + 1.2 * envelope * sin(2 * pi * 6 * t_group)
    trial_list[[row_i]] <- data.frame(
      subject = paste0("s", subject_i),
      trial_index = seq_len(n_bins) + (rep_i - 1L) * n_bins,
      time_ms = nominal_bins + stats::runif(n_bins, -8, 8),
      score = true_signal + 0.003 * seq_len(n_bins) + stats::rnorm(n_bins, sd = 0.25)
    )
  }
}
grouped_trials <- do.call(rbind, trial_list)
```

## Your first oscillation score in 30 seconds

Create a simple 10 Hz event train, compute the score, and test significance
all in one call. This example is intentionally clean so the detected peak
lands at 10 Hz, making the basic workflow easy to interpret:

```{r first-score}
fs <- 1024
sig <- numeric(2048)
spikes <- seq(52, 1996, by = 102)
sig[spikes] <- 1

result <- oscillation_score_z(sig, fs = fs, flim = c(5, 20), nrep = 200)

stopifnot(
  is.finite(result$oscore),
  result$oscore > 0,
  is.finite(result$fosc),
  result$fosc >= 5,
  result$fosc <= 20,
  is.finite(result$z),
  is.finite(result$pval),
  result$pval >= 0,
  result$pval <= 1
)
```

```{r show-result}
cat("Oscillation score:", round(result$oscore, 3), "\n")
cat("Peak frequency:   ", round(result$fosc, 2), "Hz\n")
cat("Z-score:          ", round(result$z, 2), "\n")
cat("P-value:          ", format(result$pval, digits = 4), "\n")
cat("Significant:      ", result$significant, "\n")
```

```{r plot-first-spectrum, echo=FALSE, fig.cap="Power spectrum of the autocorrelogram. The dashed line marks the detected peak at 10 Hz.", fig.height=3}
sp <- spectral_peak(sig, fs = fs, flim = c(1, 50))
plot_spectrum(sp$fxx, sp$spectrum, peak = result$fosc)
```

The oscillation score is `r round(result$oscore, 2)` with a z-score of
`r round(result$z, 1)` — clear evidence for rhythmic structure in the 5-20 Hz
target band. Because this toy signal is perfectly regular, the detected peak
lands exactly at 10 Hz. Real behavioral data are usually noisier, so small
departures from the nominal frequency are normal in practice.

## Extracting phase at behavioral events

`phase_at_events()` does not need a special object. It just needs a numeric
vector of event times on a common time axis, usually one RT per trial.

In a behavioral study, that usually means the latency from a common alignment
event, such as cue onset or stimulus onset, to the response on each trial.
If your RTs already live in a vector, use that vector directly. If your data
are in a trial-wise data frame with one row per trial, extract the RT column
after subsetting to the trials you want to analyze. If the RTs are stored in
milliseconds, convert them to seconds first, because both `events` and `dt`
are in seconds. If your table stores absolute timestamps instead of latencies,
subtract the alignment time first so each trial contributes a comparable RT.

Here are the two most common setups:

```{r event-inputs}
rt_ms <- c(412, 438, 521, 547, 603, 618)

trial_data <- data.frame(
  trial = seq_along(rt_ms),
  correct = c(TRUE, TRUE, TRUE, FALSE, TRUE, TRUE),
  rt_ms = rt_ms
)

events_from_vector <- rt_ms / 1000
events_from_df <- subset(trial_data, correct)$rt_ms / 1000

stopifnot(
  is.numeric(events_from_vector),
  all(events_from_vector > 0),
  is.numeric(events_from_df),
  all(events_from_df > 0)
)

events_from_df
```

If you have multiple subjects or conditions, do this extraction separately for
each subset rather than pooling everything into one vector. Once you have a
numeric RT vector in seconds, the call is the same. Here `dt = 0.001` means a
1 ms sampling grid for the continuous trace:

```{r phase-events}
events <- c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85, 0.95)
phases <- phase_at_events(events, dt = 0.001, fosc = 10, bandwidth = 1)

stopifnot(
  length(phases) == length(events),
  all(is.finite(phases)),
  all(phases >= -pi),
  all(phases <= pi)
)
```

```{r plot-phases, echo=FALSE, fig.cap="Phase distribution at event times.", fig.height=3}
plot_phase_hist(phases)
```

Test whether the phases are non-uniformly distributed:

```{r rayleigh}
ray <- circ_rayleigh(phases)
r <- circ_r(phases)
stopifnot(is.finite(r), r >= 0, r <= 1, is.finite(ray$pval), ray$pval >= 0, ray$pval <= 1)

cat("Mean resultant length:", round(r, 3), "\n")
cat("Rayleigh p-value:     ", format(ray$pval, digits = 3), "\n")
```

## How do you move from trials to grouped oscillation maps?

When you have many subjects, the grouped workflow is:

1. align each trial to a target bin grid
2. regress out nuisance structure
3. aggregate to one value per subject and bin
4. compute grouped spectral, time-frequency, or phase summaries

The example below uses a cue-locked behavioral trace with small timing jitter
and a slow linear drift. The true rhythmic component is centered near 6 Hz.
This dense-sampling trial-to-spectrum pattern is also the kind of workflow used
in recent empirical applications of cue-locked memory analysis. For a
reproducible example based on the shared Biba et al. (2026) dataset, see
`inst/examples/biba_confirmatory_fft.R`.

```{r aggregate-trials}
aligned_trials <- align_time_bins(
  grouped_trials,
  time = "time_ms",
  bins = nominal_bins,
  tolerance = 18,
  out = "time_bin"
)

residual_trials <- residualize_trials(
  aligned_trials,
  response = "score",
  terms = "trial_index",
  by = "subject",
  out = "score_resid"
)

grouped_bins <- aggregate_by_bin(
  residual_trials,
  value = "score_resid",
  bin = "time_bin",
  by = "subject",
  bins = nominal_bins
)

group_mat <- unname(as.matrix(stats::xtabs(score_resid ~ subject + time_bin, data = grouped_bins)))

stopifnot(
  nrow(group_mat) == n_subjects,
  ncol(group_mat) == length(nominal_bins),
  all(is.finite(group_mat))
)

dim(group_mat)
```

### What is the dominant grouped rhythm?

```{r grouped-spectrum}
spec <- group_spectrum_test(
  group_mat,
  fs = fs_group,
  flim = c(3, 10),
  null = "shuffle_labels",
  nrep = 100,
  taper = "hann",
  seed = 11
)

peak_i <- which.max(spec$observed$power)

stopifnot(
  is.finite(spec$observed$peak),
  abs(spec$observed$peak - 6) < 1,
  is.finite(spec$stats$z[peak_i]),
  spec$stats$p_adj[peak_i] < 0.1
)

cat("Grouped peak frequency:", round(spec$observed$peak, 2), "Hz\n")
cat("Peak adjusted p-value:", format(spec$stats$p_adj[peak_i], digits = 3), "\n")
cat("Peak z-score:         ", round(spec$stats$z[peak_i], 2), "\n")
```

```{r plot-grouped-spectrum, echo=FALSE, fig.cap="Grouped spectrum after trial alignment, nuisance regression, and bin aggregation.", fig.height=3}
plot_spectrum(spec$observed$freq, spec$observed$power, peak = spec$observed$peak)
```

### Where is the effect in time-frequency space?

```{r grouped-tfr}
tfr <- group_tfr_test(
  group_mat,
  fs = fs_group,
  freqs = 4:8,
  bandwidth = 0.75,
  null = "shuffle_labels",
  nrep = 40,
  reflect = TRUE,
  edge = 4,
  seed = 11
)

peak_cell <- which(tfr$observed$map == max(tfr$observed$map), arr.ind = TRUE)[1, ]
peak_freq <- tfr$observed$freq[peak_cell[1]]
peak_time <- tfr$observed$time[peak_cell[2]]

stopifnot(
  is.finite(tfr$stats$z[peak_cell[1], peak_cell[2]]),
  peak_freq >= 5,
  peak_freq <= 7,
  peak_time >= min(tfr$observed$time),
  peak_time <= max(tfr$observed$time)
)

cat("Max TFR frequency:", peak_freq, "Hz\n")
cat("Max TFR time:     ", round(peak_time), "ms\n")
cat("Max TFR z-score:  ", round(tfr$stats$z[peak_cell[1], peak_cell[2]], 2), "\n")
cat("Detected clusters:", length(tfr$stats$clusters), "\n")
```

```{r plot-grouped-tfr, echo=FALSE, fig.cap="Grouped time-frequency z-map. Cluster outlines mark suprathreshold regions.", fig.height=4}
plot_cluster_heatmap(
  tfr$stats$z,
  clusters = tfr$stats$clusters,
  freq_axis = tfr$observed$freq,
  time_axis = tfr$observed$time
)
```

### Are phases aligned across subjects?

```{r grouped-phase}
phase_grp <- group_phase_consistency(
  group_mat,
  fs = fs_group,
  freq = 6,
  bandwidth = 0.75,
  reflect = TRUE,
  edge = 4
)

stopifnot(
  all(is.finite(phase_grp$stats$r)),
  all(phase_grp$stats$r >= 0),
  all(phase_grp$stats$r <= 1),
  mean(phase_grp$stats$r) > 0.6
)

cat("Mean phase concentration:", round(mean(phase_grp$stats$r), 3), "\n")
cat("Peak phase concentration:", round(max(phase_grp$stats$r), 3), "\n")
```

## Where to go next

You've just seen the core three-step pipeline: **score** (`oscillation_score_z()`),
**extract phase** (`phase_at_events()`), **test** (`circ_rayleigh()`), and the
grouped workflow from trial tables to spectral maps. When you want to go
deeper, the other vignettes unpack each layer in more detail.

| I want to... | Read this |
|:-------------|:----------|
| Understand oscillation scores in depth | `vignette("oscillation-score")` |
| Analyze phase consistency and clusters | `vignette("ppc-clustering")` |
| Run batch analyses across subjects | `vignette("workflow-wrappers")` |

## References

**bosc** is an R port of the
[behavioral-oscillations](https://github.com/marijeterwal/behavioral-oscillations)
MATLAB toolbox. If you use this package in published work, please cite the
original paper:

> Ter Wal, M., Linde-Domingo, J., Lifanov, J., Roux, F., Kolibius, L.D.,
> Gollwitzer, S., Lang, J., Hamer, H., Rollings, D., Sawlani, V.,
> Chelvarajah, R., Staresina, B., Hanslmayr, S., & Wimber, M. (2021).
> Theta rhythmicity governs the timing of behavioural and hippocampal
> responses in humans specifically during memory-dependent tasks.
> *Nature Communications*, 12, 7048.
> <https://doi.org/10.1038/s41467-021-27323-3>

Related dense-sampling applications include:

> Biba, T. M., Decker, A., Herrmann, B., Fukuda, K., Katz, C. N.,
> Valiante, T. A., & Duncan, K. (2026). Episodic memory encoding fluctuates
> at a theta rhythm of 3-10 Hz. *Nature Human Behaviour*.
> <https://doi.org/10.1038/s41562-026-02416-5>
