---
title: Get started with bosc
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with bosc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

## What does bosc do?

You have behavioral time series — reaction times, button presses, spike
trains — and you want to know whether they contain rhythmic structure. Is
there a dominant oscillation frequency? Is it statistically significant? At
what phase of the oscillation do events occur?

**bosc** answers these questions with a small, focused API:

| Verb | Function | What it does |
|:-----|:---------|:-------------|
| **Score** | `oscillation_score_z()` | Quantify oscillation strength + significance |
| **Filter** | `narrowband_hilbert()` | Extract phase and amplitude at a frequency |
| **Phase** | `phase_at_events()` | Get instantaneous phase at event times |
| **Cluster** | `detect_clusters()` | Find significant regions in time-frequency maps |
| **Test** | `circ_rayleigh()` | Test for non-uniform phase distributions |

```{r load-bosc}
library(bosc)
```

## Your first oscillation score in 30 seconds

Create a signal with 10 Hz periodicity, compute the score, and test
significance — all in one call:

```{r first-score}
set.seed(42)
fs <- 1000
sig <- numeric(2000)
spikes <- seq(50, 1950, by = 100) + round(rnorm(20, sd = 5))
spikes <- spikes[spikes > 0 & spikes <= 2000]
sig[spikes] <- 1

result <- oscillation_score_z(sig, fs = fs, flim = c(5, 20), nrep = 200)
```

```{r show-result}
cat("Oscillation score:", round(result$oscore, 3), "\n")
cat("Peak frequency:   ", round(result$fosc, 2), "Hz\n")
cat("Z-score:          ", round(result$z, 2), "\n")
cat("Significant:      ", result$significant, "\n")
```

```{r plot-first-spectrum, echo=FALSE, fig.cap="Power spectrum of the autocorrelogram. The dashed line marks the detected peak at ~10 Hz.", fig.height=3}
sp <- spectral_peak(sig, fs = fs, flim = c(1, 50))
plot_spectrum(sp$fxx, sp$spectrum, peak = result$fosc)
```

The oscillation score is `r round(result$oscore, 2)` with a z-score of
`r round(result$z, 1)` — a clear 10 Hz rhythm, as expected.

## Extracting phase at behavioral events

Given event times (in seconds), `phase_at_events()` builds a continuous
trace, filters at the target frequency, and returns the instantaneous phase
at each event:

```{r phase-events}
events <- c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85, 0.95)
phases <- phase_at_events(events, dt = 0.001, fosc = 10, bandwidth = 1)
```

```{r plot-phases, echo=FALSE, fig.cap="Phase distribution at event times.", fig.height=3}
plot_phase_hist(phases)
```

Test whether the phases are non-uniformly distributed:

```{r rayleigh}
ray <- circ_rayleigh(phases)
cat("Mean resultant length:", round(circ_r(phases), 3), "\n")
cat("Rayleigh p-value:     ", format(ray$pval, digits = 3), "\n")
```

## Where to go next

You've just seen the core three-step pipeline: **score** (`oscillation_score_z()`),
**extract phase** (`phase_at_events()`), **test** (`circ_rayleigh()`). For
group-level analyses, add `paired_tscore()` and `detect_clusters()` for
cluster-based permutation testing on time-frequency maps.

| I want to... | Read this |
|:-------------|:----------|
| Understand oscillation scores in depth | `vignette("oscillation-score")` |
| Analyze phase consistency and clusters | `vignette("ppc-clustering")` |
| Run batch analyses across subjects | `vignette("workflow-wrappers")` |

## References

**bosc** is an R port of the
[behavioral-oscillations](https://github.com/marijeterwal/behavioral-oscillations)
MATLAB toolbox. If you use this package in published work, please cite the
original paper:

> Ter Wal, M., Linde-Domingo, J., Lifanov, J., Roux, F., Kolibius, L.D.,
> Gollwitzer, S., Lang, J., Hamer, H., Rollings, D., Sawlani, V.,
> Chelvarajah, R., Staresina, B., Hanslmayr, S., & Wimber, M. (2021).
> Theta rhythmicity governs the timing of behavioural and hippocampal
> responses in humans specifically during memory-dependent tasks.
> *Nature Communications*, 12, 7048.
> <https://doi.org/10.1038/s41467-021-25959-7>
