---
title: "Oscillation Score Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oscillation Score Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

The **oscillation score** quantifies the strength of periodic (rhythmic) structure
in behavioral data such as spike trains, button presses, or continuous response
signals. It was developed to detect behavioral oscillations in tasks where
subjects respond rhythmically, potentially phase-locked to neural oscillations.

This vignette demonstrates:

- Computing oscillation scores from spike trains
- Surrogate testing for statistical significance
- Multi-band analysis
- Narrowband filtering for phase extraction

For phase consistency analysis and cluster detection, see `vignette("ppc-clustering")`.

## Setup

```{r load-library}
library(bosc)
```

## Synthetic Spike Train Example

We generate a spike train with 10 Hz periodicity embedded in noise to demonstrate
the oscillation score computation.
```{r generate-data}
set.seed(42)
fs <- 1000  # 1 kHz sampling rate
duration <- 5  # 5 seconds
n_samples <- fs * duration

# Create periodic spike train (10 Hz = every 100 samples)
sig <- numeric(n_samples)
spike_times <- seq(50, n_samples - 50, by = 100)
# Add jitter to make it realistic
spike_times <- spike_times + round(rnorm(length(spike_times), sd = 5))
spike_times <- spike_times[spike_times > 0 & spike_times <= n_samples]
sig[spike_times] <- 1

# Add some random spikes as noise
n_noise <- 50
noise_times <- sample(setdiff(seq_len(n_samples), spike_times), n_noise)
sig[noise_times] <- 1

cat("Total spikes:", sum(sig), "\n")
cat("Periodic spikes:", length(spike_times), "\n")
```

## Computing the Oscillation Score

The `oscillation_score()` function computes the oscillation score through
these steps:

1. Computing the autocorrelogram of the signal
2. Removing the central peak (to focus on periodicity, not spike width)
3. Finding the spectral peak in the frequency band of interest
4. Returning the ratio of peak power to mean power

```{r compute-oscore}
# Compute oscillation score in the 5-20 Hz band
result <- oscillation_score(
  signal = sig,
  fs = fs,
  flim = c(5, 20),
  smoothach = TRUE,
  warnings = TRUE
)

cat("Oscillation Score:", round(result$oscore, 3), "\n")
cat("Peak Frequency:", round(result$fosc, 2), "Hz\n")
cat("Effective frequency limits:", round(result$flim, 2), "Hz\n")
```

## Visualizing the Spectrum

```{r plot-spectrum, fig.cap="Power spectrum of the autocorrelogram showing peak at ~10 Hz"}
# Get spectrum for visualization
sp <- spectral_peak(sig, fs = fs, flim = c(1, 50))

if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_spectrum(sp$fxx, sp$spectrum, peak = result$fosc)
}
```

## Surrogate Testing

To assess statistical significance, we compare the observed oscillation score
against a null distribution generated by shuffling spike times. This preserves
basic signal properties (spike count, duration) while destroying temporal
structure.

```{r surrogate-test}
# Generate surrogate distribution (100 shuffles)
surrogates <- oscillation_score_surrogates(
  signal = sig,
  fs = fs,
  flim = c(5, 20),
  nrep = 100,
  fpeak = result$fosc,
  keep_trend = FALSE,
  warnings = FALSE
)

# Compute z-score against null distribution
mean_sur <- mean(surrogates$oscore_rp, na.rm = TRUE)
sd_sur <- sd(surrogates$oscore_rp, na.rm = TRUE)
z_score <- (result$oscore - mean_sur) / sd_sur

cat("Surrogate mean:", round(mean_sur, 3), "\n")
cat("Surrogate SD:", round(sd_sur, 3), "\n")
cat("Z-score:", round(z_score, 2), "\n")
cat("P-value (one-tailed):", round(pnorm(-abs(z_score)), 4), "\n")
```

## Comparing Different Frequency Bands

A common analysis examines oscillation scores across canonical frequency bands
to identify the dominant rhythm:

```{r multi-band}
bands <- list(
  theta = c(4, 8),
  alpha = c(8, 12),
  beta = c(12, 30)
)

results <- lapply(bands, function(flim) {
  res <- oscillation_score(sig, fs = fs, flim = flim, warnings = FALSE)
  c(oscore = res$oscore, fosc = res$fosc)
})

results_df <- do.call(rbind, results)
print(round(results_df, 3))
```

As expected, the alpha band (8-12 Hz) shows the highest oscillation score since
our synthetic signal has 10 Hz periodicity.

## Narrowband Hilbert Analysis

To extract instantaneous phase and amplitude at a specific frequency, use
`narrowband_hilbert()`. This is essential for phase-locking analyses.

```{r narrowband-hilbert}
# Create a continuous oscillatory signal for demonstration
t <- seq(0, 2, by = 1/fs)
continuous_sig <- sin(2 * pi * 10 * t) + 0.5 * sin(2 * pi * 25 * t) + rnorm(length(t), sd = 0.3)

# Extract 10 Hz component (alpha band: 8-12 Hz)
nb <- narrowband_hilbert(
  data = continuous_sig,
  fs = fs,
  freqlim = c(8, 12)
)

cat("Analytic signal length:", length(nb$analytic), "\n")
cat("Mean amplitude:", round(mean(Mod(nb$analytic)), 3), "\n")

# Phase extraction
phases <- Arg(nb$analytic)
cat("Phase range:", round(range(phases), 2), "radians\n")
```

## Using make_continuous_trace

If your data consists of event times (in seconds) rather than a binary signal,
use `make_continuous_trace()` to convert to a continuous time series:

```{r continuous-trace}
# Event times in seconds
event_times <- spike_times / fs

# Convert to continuous trace with Gaussian smoothing
trace <- make_continuous_trace(
  events = event_times,
  dt = 0.001,  # 1 ms bins
  sd_smooth = 0.005  # 5 ms Gaussian smoothing
)

cat("Trace length:", length(trace$signal), "samples\n")
cat("Time span:", round(range(trace$tspan), 2), "seconds\n")

# Can now compute oscillation score on the smoothed trace
trace_result <- oscillation_score(
  signal = trace$signal,
  fs = 1 / 0.001,  # fs from dt

  flim = c(5, 20),
  warnings = FALSE
)
cat("Oscillation score (smoothed):", round(trace_result$oscore, 3), "\n")
```

## Summary

Key functions demonstrated:

| Function | Purpose |
|----------|---------|
| `oscillation_score()` | Compute oscillation score from signal |
| `oscillation_score_surrogates()` | Generate null distribution for significance testing |
| `spectral_peak()` | Find peak frequency in power spectrum |
| `narrowband_hilbert()` | Extract phase/amplitude at specific frequency |
| `make_continuous_trace()` | Convert event times to continuous signal |
| `plot_spectrum()` | Visualize power spectrum |

## Session Info

```{r session-info}
sessionInfo()
```
