---
title: Oscillation Score Analysis
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oscillation Score Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

## Why oscillation scores?

You've recorded behavioral responses — button presses, saccades, or reaction
times — and you suspect they have a rhythmic structure. Maybe they cluster at a
particular frequency, phase-locked to an underlying neural oscillation. How do
you quantify that rhythm and test whether it's real?

The **oscillation score** answers this by measuring the strength of periodic
structure in your time series. It works by computing an autocorrelogram,
removing the central peak, and finding the spectral peak in the frequency band
of interest. This vignette walks you through the full pipeline: computing
scores, testing significance with surrogates, scanning across frequency bands,
and extracting instantaneous phase.

## Setup

```{r load-library}
library(bosc)
```

## Synthetic Spike Train Example

Let's create a spike train with 10 Hz periodicity embedded in noise to demonstrate
the oscillation score computation.

```{r generate-data}
set.seed(42)
fs <- 1000  # 1 kHz sampling rate
duration <- 5  # 5 seconds
n_samples <- fs * duration

# Create periodic spike train (10 Hz = every 100 samples)
sig <- numeric(n_samples)
spike_times <- seq(50, n_samples - 50, by = 100)
# Add jitter to make it realistic
spike_times <- spike_times + round(rnorm(length(spike_times), sd = 5))
spike_times <- spike_times[spike_times > 0 & spike_times <= n_samples]
sig[spike_times] <- 1

# Add some random spikes as noise
n_noise <- 50
noise_times <- sample(setdiff(seq_len(n_samples), spike_times), n_noise)
sig[noise_times] <- 1

cat("Total spikes:", sum(sig), "\n")
cat("Periodic spikes:", length(spike_times), "\n")
```

Here's what our synthetic signal looks like — periodic spikes at 10 Hz with some
jitter and noise spikes mixed in:

```{r plot-spikes, echo=FALSE, fig.cap="Synthetic spike train with 10 Hz periodicity.", fig.height=2.5}
plot(seq_len(n_samples) / fs, sig, type = "h", col = "#1B75BB",
     xlab = "Time (s)", ylab = "Spike", main = "",
     lwd = 0.5, las = 1)
```

## Computing the Oscillation Score

The `oscillation_score()` function computes the oscillation score through
these steps:

1. Computing the autocorrelogram of the signal
2. Removing the central peak (to focus on periodicity, not spike width)
3. Finding the spectral peak in the frequency band of interest
4. Returning the ratio of peak power to mean power

```{r compute-oscore}
# Compute oscillation score in the 5-20 Hz band
result <- oscillation_score(
  signal = sig,
  fs = fs,
  flim = c(5, 20),
  smoothach = TRUE,
  warnings = TRUE
)

cat("Oscillation Score:", round(result$oscore, 3), "\n")
cat("Peak Frequency:", round(result$fosc, 2), "Hz\n")
cat("Effective frequency limits:", round(result$flim, 2), "Hz\n")
```

## Visualizing the Spectrum

```{r plot-spectrum, fig.cap="Power spectrum of the autocorrelogram showing peak at ~10 Hz"}
# Get spectrum for visualization
sp <- spectral_peak(sig, fs = fs, flim = c(1, 50))

plot_spectrum(sp$fxx, sp$spectrum, peak = result$fosc)
```

## Surrogate Testing

To assess statistical significance, we compare the observed oscillation score
against a null distribution generated by shuffling spike times. This preserves
basic signal properties (spike count, duration) while destroying temporal
structure.

```{r surrogate-test}
# Generate surrogate distribution (100 shuffles)
surrogates <- oscillation_score_surrogates(
  signal = sig,
  fs = fs,
  flim = c(5, 20),
  nrep = 100,
  fpeak = result$fosc,
  keep_trend = FALSE,
  warnings = FALSE
)

# Compute z-score against null distribution
mean_sur <- mean(surrogates$oscore_rp, na.rm = TRUE)
sd_sur <- sd(surrogates$oscore_rp, na.rm = TRUE)
z_score <- (result$oscore - mean_sur) / sd_sur

cat("Surrogate mean:", round(mean_sur, 3), "\n")
cat("Surrogate SD:", round(sd_sur, 3), "\n")
cat("Z-score:", round(z_score, 2), "\n")
cat("P-value (one-tailed):", round(pnorm(-abs(z_score)), 4), "\n")
```

```{r plot-surrogates, echo=FALSE, fig.cap="Observed oscillation score (red line) against the surrogate null distribution.", fig.height=3}
xlim_range <- range(c(surrogates$oscore_rp, result$oscore), na.rm = TRUE)
xlim_range <- xlim_range + c(-1, 1) * diff(xlim_range) * 0.05
hist(surrogates$oscore_rp, breaks = 20, col = "grey80", border = "white",
     main = "", xlab = "Oscillation Score", ylab = "Count", las = 1,
     xlim = xlim_range)
abline(v = result$oscore, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("Observed", "Surrogates"),
       col = c("red", "grey60"), lwd = c(2, 10), lty = c(2, 1), bty = "n")
```

## Comparing Different Frequency Bands

A common analysis examines oscillation scores across canonical frequency bands
to identify the dominant rhythm:

```{r multi-band}
bands <- list(
  theta = c(4, 8),
  alpha = c(8, 12),
  beta = c(12, 30)
)

results <- lapply(names(bands), function(band_name) {
  flim <- bands[[band_name]]
  res <- oscillation_score(sig, fs = fs, flim = flim, warnings = FALSE)
  data.frame(
    band = band_name,
    req_fmin = flim[1],
    req_fmax = flim[2],
    eff_fmin = if (length(res$flim) == 2) res$flim[1] else NA_real_,
    eff_fmax = if (length(res$flim) == 2) res$flim[2] else NA_real_,
    oscore = res$oscore,
    fosc = res$fosc,
    stringsAsFactors = FALSE
  )
})

results_df <- do.call(rbind, results)
results_df_print <- results_df
num_cols <- vapply(results_df_print, is.numeric, logical(1))
results_df_print[num_cols] <- lapply(results_df_print[num_cols], round, 3)
print(results_df_print)
```

As expected, the alpha band (8-12 Hz) shows the highest oscillation score since
our synthetic signal has 10 Hz periodicity.

`beta` can legitimately return `NA` for `oscore`/`fosc` here. Internally,
`oscillation_score()` constrains the requested band to an effective range based
on the data support:

`fmax <- min(flim[2], sum(sigq) / (length(sigq) / fs))`

For this sparse spike train, the effective beta range is narrower than
12-30 Hz, and no stable local spectral peak is found in that range. In that
case, the function returns `NA`, which indicates "no detectable beta-band peak"
rather than an error.

```{r plot-bands, echo=FALSE, fig.cap="Oscillation scores across frequency bands. The alpha band (containing our 10 Hz signal) shows the strongest score.", fig.height=3}
scores <- setNames(results_df$oscore, results_df$band)
barplot(scores, names.arg = names(scores), col = "#1B75BB",
        ylab = "Oscillation Score", las = 1, border = NA)
```

## Narrowband Hilbert Analysis

To extract instantaneous phase and amplitude at a specific frequency, use
`narrowband_hilbert()`. This is essential for phase-locking analyses.

```{r narrowband-hilbert}
# Create a continuous oscillatory signal for demonstration
t <- seq(0, 2, by = 1/fs)
continuous_sig <- sin(2 * pi * 10 * t) + 0.5 * sin(2 * pi * 25 * t) + rnorm(length(t), sd = 0.3)

# Extract 10 Hz component (alpha band: 8-12 Hz)
nb <- narrowband_hilbert(
  data = continuous_sig,
  fs = fs,
  freqlim = c(8, 12)
)

cat("Analytic signal length:", length(nb$analytic), "\n")
cat("Mean amplitude:", round(mean(Mod(nb$analytic)), 3), "\n")

# Phase extraction
phases <- Arg(nb$analytic)
cat("Phase range:", round(range(phases), 2), "radians\n")
```

```{r plot-hilbert, echo=FALSE, fig.cap="Narrowband Hilbert decomposition: raw signal (top), filtered 8-12 Hz component (middle), and instantaneous phase (bottom).", fig.height=5}
t_plot <- t[1:500]
par(mfrow = c(3, 1), mar = c(3, 4, 1.5, 1))
plot(t_plot, continuous_sig[1:500], type = "l", col = "grey50",
     xlab = "", ylab = "Amplitude", main = "Raw signal", las = 1)
plot(t_plot, nb$filtered[1:500], type = "l", col = "#1B75BB",
     xlab = "", ylab = "Amplitude", main = "Filtered (8-12 Hz)", las = 1)
plot(t_plot, phases[1:500], type = "l", col = "#D35400",
     xlab = "Time (s)", ylab = "Phase (rad)", main = "Instantaneous phase", las = 1)
abline(h = c(-pi, 0, pi), col = "grey80", lty = 3)
```

## Using make_continuous_trace

If your data consists of event times (in seconds) rather than a binary signal,
use `make_continuous_trace()` to convert to a continuous time series:

```{r continuous-trace}
# Event times in seconds
event_times <- spike_times / fs

# Convert to continuous trace with Gaussian smoothing
trace <- make_continuous_trace(
  events = event_times,
  dt = 0.001,  # 1 ms bins
  sd_smooth = 0.005  # 5 ms Gaussian smoothing
)

cat("Trace length:", length(trace$signal), "samples\n")
cat("Time span:", round(range(trace$tspan), 2), "seconds\n")

# Can now compute oscillation score on the smoothed trace
trace_result <- oscillation_score(
  signal = trace$signal,
  fs = 1 / 0.001,  # fs from dt
  flim = c(5, 20),
  warnings = FALSE
)
cat("Oscillation score (smoothed):", round(trace_result$oscore, 3), "\n")
```

## Next steps

- Test significance with config-driven batch workflows: `vignette("workflow-wrappers")`
- Analyze phase consistency across trials: `vignette("ppc-clustering")`
- See `?oscillation_score` for the full argument list

## References

The algorithms in **bosc** are ported from the
[behavioral-oscillations](https://github.com/marijeterwal/behavioral-oscillations)
MATLAB toolbox. Please cite:

> Ter Wal, M. et al. (2021). Theta rhythmicity governs the timing of
> behavioural and hippocampal responses in humans specifically during
> memory-dependent tasks. *Nature Communications*, 12, 7048.
> <https://doi.org/10.1038/s41467-021-25959-7>
