---
title: PPC Analysis and Cluster Detection
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PPC Analysis and Cluster Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
has_imager <- requireNamespace("imager", quietly = TRUE)
```

## The problem: finding rhythmic structure in noisy data

You've collected phase angles from neural recordings across many trials —
perhaps from narrowband-filtered EEG time-locked to a behavioral response.
Two questions arise:

1. **Are the phases consistent across trials?** If subjects respond at a
   preferred oscillatory phase, the phase distribution will be non-uniform.
2. **Where in time-frequency space is the effect?** With data at many
   frequencies and time points, you need cluster-based correction to control
   for multiple comparisons.

This vignette covers both: **Pairwise Phase Consistency (PPC)** for
bias-free phase-locking measurement, and **cluster-based permutation
testing** for time-frequency maps. PPC avoids the trial-count bias of mean
resultant length, making it the preferred metric for comparing conditions
with unequal trial numbers.

For oscillation score analysis, see `vignette("oscillation-score")`.

## Setup

```{r load-library}
library(bosc)
```

## Pairwise Phase Consistency (PPC)

### What is PPC?
PPC measures how consistent phase angles are across trials by computing the
average cosine of pairwise phase differences:

$$\text{PPC} = \frac{2}{n(n-1)} \sum_{i<j} \cos(\phi_i - \phi_j)$$

- PPC = 1: All phases identical (perfect consistency)
- PPC = 0: Random phases (no consistency)
- PPC < 0: Phases clustered at opposite angles

### Basic PPC Computation

```{r ppc-basic}
set.seed(123)

# Simulate phase data: 20 trials, 10 time points
n_trials <- 20

# Condition 1: Phases clustered around 0 (high consistency)
phases_consistent <- matrix(
  rnorm(n_trials * 10, mean = 0, sd = 0.3),
  nrow = n_trials
)

# Condition 2: Random phases (low consistency)
phases_random <- matrix(
  runif(n_trials * 10, -pi, pi),
  nrow = n_trials
)

# Compute PPC along trial dimension (dim = 1)
ppc_consistent <- pairwise_phase_consistency(phases_consistent, dim = 1)
ppc_random <- pairwise_phase_consistency(phases_random, dim = 1)

cat("PPC (consistent phases):", round(mean(ppc_consistent), 3), "\n")
cat("PPC (random phases):", round(mean(ppc_random), 3), "\n")
```

```{r plot-ppc-comparison, echo=FALSE, fig.cap="PPC values across time points: consistent phases (blue) vs random phases (grey).", fig.height=3}
par(mar = c(4, 4, 1, 1))
plot(seq_along(ppc_consistent), ppc_consistent, type = "b", pch = 19,
     col = "#1B75BB", ylim = range(c(ppc_consistent, ppc_random)),
     xlab = "Time point", ylab = "PPC", las = 1)
lines(seq_along(ppc_random), ppc_random, type = "b", pch = 19, col = "grey60")
abline(h = 0, lty = 3, col = "grey80")
legend("topright", legend = c("Consistent", "Random"),
       col = c("#1B75BB", "grey60"), pch = 19, lty = 1, bty = "n")
```

### Phase Extraction with Narrowband Hilbert

In real analyses, phases come from narrowband-filtered neural signals:

```{r phase-extraction}
set.seed(456)
fs <- 500  # 500 Hz sampling

# Simulate 20 trials of oscillatory data with phase-locked response
n_trials <- 20
trial_length <- fs  # 1 second per trial

# Generate trials with consistent phase at response time
phases_at_response <- numeric(n_trials)
for (i in seq_len(n_trials)) {
  # Simulated neural signal: 10 Hz oscillation + noise
  t <- seq(0, 1, length.out = trial_length)
  neural <- sin(2 * pi * 10 * t + runif(1, -0.3, 0.3)) + rnorm(trial_length, sd = 0.5)

  # Extract alpha phase using narrowband Hilbert
  nb <- narrowband_hilbert(data = neural, fs = fs, freqlim = c(8, 12))

  # Get phase at "response time" (sample 250 = 500 ms)
  phases_at_response[i] <- Arg(nb$analytic[250])
}

# Compute circular statistics
cat("Mean phase:", round(circ_mean(phases_at_response), 3), "rad\n")
cat("Resultant length:", round(circ_r(phases_at_response), 3), "\n")

# Rayleigh test
ray <- circ_rayleigh(phases_at_response)
cat("Rayleigh p-value:", format(ray$pval, digits = 3), "\n")
```

```{r plot-phase-rose, echo=FALSE, fig.cap="Phase angles at response time across 20 trials. Clustering near 0 indicates phase-locking.", fig.height=3.5, fig.width=3.5}
par(mar = c(1, 1, 1, 1))
plot(cos(phases_at_response), sin(phases_at_response), pch = 19,
     col = "#1B75BB", asp = 1, xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3),
     xlab = "", ylab = "", axes = FALSE)
# Draw unit circle
theta_seq <- seq(0, 2 * pi, length.out = 100)
lines(cos(theta_seq), sin(theta_seq), col = "grey70")
lines(c(0, 0), c(-1.1, 1.1), col = "grey85", lty = 3)
lines(c(-1.1, 1.1), c(0, 0), col = "grey85", lty = 3)
# Mean direction arrow
r <- circ_r(phases_at_response)
mu <- circ_mean(phases_at_response)
arrows(0, 0, r * cos(mu), r * sin(mu), col = "red", lwd = 2, length = 0.1)
```

### Circular Statistics Functions

```{r circular-stats}
# Using the phases from consistent condition
angles <- phases_consistent[, 1]

# Circular mean direction
mu <- circ_mean(angles)
cat("Mean direction:", round(mu, 3), "rad\n")

# Mean resultant length (concentration)
r <- circ_r(angles)
cat("Resultant length:", round(r, 3), "\n")

# Rayleigh test for non-uniformity
rayleigh <- circ_rayleigh(angles)
cat("Rayleigh p-value:", format(rayleigh$pval, digits = 3), "\n")

# V-test against expected direction of 0
vtest <- circ_vtest(angles, dir = 0)
cat("V-test p-value:", format(vtest$pval, digits = 3), "\n")
```

### Visualizing Phase Distribution

```{r phase-histogram, fig.cap="Phase distribution showing clustering around 0"}
plot_phase_hist(angles, nbins = 16)
```

## How do you find significant clusters in time-frequency maps?

Cluster-based permutation testing is the standard approach for multiple
comparison correction in time-frequency analyses. It works by:

1. Thresholding the statistical map to identify contiguous clusters
2. Computing a cluster statistic (sum of values)
3. Comparing against a null distribution from permutations

### Simulating Time-Frequency Data

We start by creating two conditions for 15 subjects across a 30-frequency ×
50-timepoint grid. The baseline condition is pure noise; we will add a
localized effect in the second step.

```{r simulate-tf-baseline}
set.seed(789)

n_freq <- 30
n_time <- 50
n_subjects <- 15

# Baseline condition: random noise across all subjects, frequencies, and times
baseline <- array(
  rnorm(n_subjects * n_freq * n_time, mean = 0, sd = 1),
  dim = c(n_subjects, n_freq, n_time)
)

# Effect condition starts identical in structure
effect <- array(
  rnorm(n_subjects * n_freq * n_time, mean = 0, sd = 1),
  dim = c(n_subjects, n_freq, n_time)
)

cat("Data dimensions:", dim(effect), "(subjects x freq x time)\n")
```

Now we inject a spatially compact effect into the theta band (frequency rows
5–10) around the response window (time points 15–25). This mimics a real
neural effect that cluster detection should recover.

```{r simulate-tf-effect}
# Add effect cluster (theta band: rows 5-10, time 15-25)
effect[, 5:10, 15:25] <- effect[, 5:10, 15:25] + 1.5

cat("Effect magnitude added to rows 5:10, times 15:25\n")
cat("Mean in effect region (effect):",
    round(mean(effect[, 5:10, 15:25]), 3), "\n")
cat("Mean in effect region (baseline):",
    round(mean(baseline[, 5:10, 15:25]), 3), "\n")
```

### Computing T-scores

```{r compute-tscores}
# Paired t-test across subjects (dimension 1)
tres <- paired_tscore(effect, baseline, dim = 1)
t_matrix <- matrix(tres$t, nrow = n_freq, ncol = n_time)

cat("T-score range:", round(range(t_matrix, na.rm = TRUE), 2), "\n")
cat("Max t-score location: freq", which.max(apply(t_matrix, 1, max)),
    ", time", which.max(apply(t_matrix, 2, max)), "\n")
```

```{r plot-tscores, echo=FALSE, fig.cap="T-score map across frequency and time. The embedded effect cluster is visible in the theta range (rows 5-10, time 15-25)."}
freq_ax_temp <- seq(2, 30, length.out = n_freq)
time_ax_temp <- seq(-0.5, 1.5, length.out = n_time)
image(time_ax_temp, freq_ax_temp, t(t_matrix), col = hcl.colors(64, "Blue-Red 3"),
      xlab = "Time (s)", ylab = "Frequency (Hz)", las = 1,
      zlim = c(-4, 4))
```

### Detecting Clusters

```{r detect-clusters, eval=has_imager}
# Define frequency and time axes
freq_axis <- seq(2, 30, length.out = n_freq)
time_axis <- seq(-0.5, 1.5, length.out = n_time)

# Detect clusters using extract method
clusters <- detect_clusters(
  data = t_matrix,
  threshold = 2.0,  # t-threshold (roughly p < 0.05 uncorrected)
  method = "extract",
  smooth = 1,
  freq_axis = freq_axis,
  time_axis = time_axis
)

cat("Number of clusters found:", length(clusters), "\n")

# Examine clusters
if (length(clusters) > 0) {
  for (i in seq_along(clusters)) {
    cl <- clusters[[i]]
    cat(sprintf("\nCluster %d:\n", i))
    cat(sprintf("  Frequency: %.1f - %.1f Hz\n", min(cl$freqs), max(cl$freqs)))
    cat(sprintf("  Time: %.2f - %.2f s\n", min(cl$times), max(cl$times)))
    cat(sprintf("  Size: %d pixels\n", length(cl$Zscores)))
    cat(sprintf("  Sum t: %.2f\n", cl$sumZscore))
    cat(sprintf("  Peak t: %.2f\n", cl$peakZscore))
  }
}
```

### Visualizing Clusters

```{r plot-clusters, eval=has_imager, fig.cap="Time-frequency map with detected clusters outlined"}
plot_cluster_heatmap(
  data = t_matrix,
  clusters = clusters,
  freq_axis = freq_axis,
  time_axis = time_axis,
  zlim = c(-4, 4)
)
```

### Watershed vs Extract Methods

Two clustering methods are available:

| Method | Description | Best for |
|--------|-------------|----------|
| `extract` | 8-connectivity connected components | Sharply defined clusters |
| `watershed` | Watershed segmentation | Smooth, overlapping regions |

```{r watershed-comparison, eval=FALSE}
# Watershed method (requires smoothing for best results)
clusters_ws <- detect_clusters(
  data = t_matrix,
  threshold = 2.0,
  method = "watershed",
  smooth = c(2, 2),  # smooth in both dimensions
  freq_axis = freq_axis,
  time_axis = time_axis
)
```

## How do you correct for multiple comparisons?

### FDR Correction

For element-wise testing without clustering, use FDR (False Discovery Rate):

```{r fdr-correction}
# Convert t-scores to p-values
df <- n_subjects - 1
p_matrix <- 2 * pt(-abs(t_matrix), df = df)

# Apply Benjamini-Hochberg FDR correction
fdr_result <- fdr_bh(as.vector(p_matrix), q = 0.05)

cat("Total tests:", length(p_matrix), "\n")
cat("Significant after FDR:", sum(fdr_result$h), "\n")
cat("Critical p-value:", format(fdr_result$crit_p, digits = 3), "\n")
```

### Non-parametric P-values

For permutation-based testing, use `nonparam_pval()`:

```{r nonparam-pval}
set.seed(111)

# Observed statistics (can be a vector)
observed_stats <- c(15.5, 12.0, 8.5)

# Null distribution from permutations
null_distribution <- rnorm(1000, mean = 10, sd = 3)

# Compute p-values
result <- nonparam_pval(observed_stats, null_distribution)
cat("Observed values:", observed_stats, "\n")
cat("P-values:", round(result$p, 4), "\n")
cat("Significant (alpha=0.05):", result$h, "\n")
```

## U-Score Matrix (Non-parametric Testing)

For comparing observed values against a permutation distribution across a matrix:

```{r uscore-example}
set.seed(222)

# Observed values (2D matrix)
observed <- matrix(rnorm(20, mean = 0.5), nrow = 4, ncol = 5)

# Reference distribution (100 permutations)
reference <- array(rnorm(100 * 4 * 5), dim = c(100, 4, 5))

# Compute U-scores (position in null distribution)
u_result <- u_score_matrix(observed, reference, alpha = 0.05)

cat("Significant elements:", sum(u_result$sgnf != 0), "of", length(observed), "\n")
```

## Putting it all together

Here's a complete workflow combining PPC analysis with cluster detection:

```{r complete-workflow, eval=has_imager}
set.seed(333)

# 1. Simulate PPC data: correct vs incorrect trials
n_subjects <- 12
n_freq <- 20
n_time <- 30

# Correct trials: higher theta PPC around response time
ppc_correct <- array(
  runif(n_subjects * n_freq * n_time, 0, 0.3),
  dim = c(n_subjects, n_freq, n_time)
)
# Add theta effect (freqs 3-6 = theta, times 10-20 = peri-response)
ppc_correct[, 3:6, 10:20] <- ppc_correct[, 3:6, 10:20] + 0.25

# Incorrect trials: baseline PPC
ppc_incorrect <- array(
  runif(n_subjects * n_freq * n_time, 0, 0.3),
  dim = c(n_subjects, n_freq, n_time)
)

# 2. Statistical comparison
diff_t <- paired_tscore(ppc_correct, ppc_incorrect, dim = 1)$t
diff_matrix <- matrix(diff_t, nrow = n_freq, ncol = n_time)

# 3. Cluster detection
freq_ax <- seq(2, 30, length.out = n_freq)
time_ax <- seq(-0.2, 0.8, length.out = n_time)

sig_clusters <- detect_clusters(
  data = diff_matrix,
  threshold = 2.0,
  method = "extract",
  smooth = 1.5,
  freq_axis = freq_ax,
  time_axis = time_ax
)

# 4. Report
cat("=== PPC: Correct vs Incorrect Trials ===\n")
cat("Subjects:", n_subjects, "\n")
cat("Clusters found:", length(sig_clusters), "\n\n")

for (i in seq_along(sig_clusters)) {
  cl <- sig_clusters[[i]]
  cat(sprintf("Cluster %d:\n", i))
  cat(sprintf("  Frequency: %.1f - %.1f Hz\n", min(cl$freqs), max(cl$freqs)))
  cat(sprintf("  Time: %.2f - %.2f s\n", min(cl$times), max(cl$times)))
  cat(sprintf("  Center: %.1f Hz, %.2f s\n", cl$CoMfreq, cl$CoMtime))
  cat(sprintf("  Cluster t-sum: %.1f\n\n", cl$sumZscore))
}
```

## Working with real data

For actual experiments, use the I/O utilities:

```{r io-example, eval=FALSE}
# Load MATLAB .mat file (requires R.matlab package)
mat_data <- read_mat("path/to/ppc_data.mat")

# Access variables
ppc_matrix <- mat_data$ppc
freq_axis <- mat_data$freqs
time_axis <- mat_data$times

# Load HDF5 file (requires hdf5r package)
h5_file <- open_hdf5("path/to/data.h5")
ppc_matrix <- h5_file[["ppc"]][]
h5_file$close_all()
```

## Next steps

- Use `oscillation_score_z()` for one-call significance testing: `vignette("workflow-wrappers")`
- Compute oscillation scores from spike trains: `vignette("oscillation-score")`
- See `?pairwise_phase_consistency` and `?detect_clusters` for full argument lists

## References

The algorithms in **bosc** are ported from the
[behavioral-oscillations](https://github.com/marijeterwal/behavioral-oscillations)
MATLAB toolbox. Please cite:

> Ter Wal, M. et al. (2021). Theta rhythmicity governs the timing of
> behavioural and hippocampal responses in humans specifically during
> memory-dependent tasks. *Nature Communications*, 12, 7048.
> <https://doi.org/10.1038/s41467-021-25959-7>
