---
title: "PPC Analysis and Cluster Detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PPC Analysis and Cluster Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates **Pairwise Phase Consistency (PPC)** analysis and
**cluster-based statistical testing** for time-frequency data. These methods
are commonly used to analyze phase-locking between behavioral responses and
neural oscillations.

**PPC** is a bias-free measure of phase consistency that, unlike mean resultant
length, does not depend on trial count. **Cluster-based testing** provides
family-wise error correction for time-frequency maps.

For oscillation score analysis, see `vignette("oscillation-score")`.

## Setup

```{r load-library}
library(bosc)
```

## Pairwise Phase Consistency (PPC)

### What is PPC?
PPC measures how consistent phase angles are across trials by computing the
average cosine of pairwise phase differences:

$$\text{PPC} = \frac{2}{n(n-1)} \sum_{i<j} \cos(\phi_i - \phi_j)$$

- PPC = 1: All phases identical (perfect consistency)
- PPC = 0: Random phases (no consistency)
- PPC < 0: Phases clustered at opposite angles

### Basic PPC Computation

```{r ppc-basic}
set.seed(123)

# Simulate phase data: 20 trials, 10 time points
n_trials <- 20

# Condition 1: Phases clustered around 0 (high consistency)
phases_consistent <- matrix(
  rnorm(n_trials * 10, mean = 0, sd = 0.3),
  nrow = n_trials
)

# Condition 2: Random phases (low consistency)
phases_random <- matrix(
  runif(n_trials * 10, -pi, pi),
  nrow = n_trials
)

# Compute PPC along trial dimension (dim = 1)
ppc_consistent <- pairwise_phase_consistency(phases_consistent, dim = 1)
ppc_random <- pairwise_phase_consistency(phases_random, dim = 1)

cat("PPC (consistent phases):", round(mean(ppc_consistent), 3), "\n")
cat("PPC (random phases):", round(mean(ppc_random), 3), "\n")
```

### Phase Extraction with Narrowband Hilbert

In real analyses, phases come from narrowband-filtered neural signals:

```{r phase-extraction}
set.seed(456)
fs <- 500  # 500 Hz sampling

# Simulate 20 trials of oscillatory data with phase-locked response
n_trials <- 20
trial_length <- fs  # 1 second per trial

# Generate trials with consistent phase at response time
phases_at_response <- numeric(n_trials)
for (i in seq_len(n_trials)) {
  # Simulated neural signal: 10 Hz oscillation + noise
  t <- seq(0, 1, length.out = trial_length)
  neural <- sin(2 * pi * 10 * t + runif(1, -0.3, 0.3)) + rnorm(trial_length, sd = 0.5)

  # Extract alpha phase using narrowband Hilbert
  nb <- narrowband_hilbert(data = neural, fs = fs, freqlim = c(8, 12))

  # Get phase at "response time" (sample 250 = 500 ms)
  phases_at_response[i] <- Arg(nb$analytic[250])
}

# Compute circular statistics
cat("Mean phase:", round(circ_mean(phases_at_response), 3), "rad\n")
cat("Resultant length:", round(circ_r(phases_at_response), 3), "\n")

# Rayleigh test
ray <- circ_rayleigh(phases_at_response)
cat("Rayleigh p-value:", format(ray$pval, digits = 3), "\n")
```

### Circular Statistics Functions

```{r circular-stats}
# Using the phases from consistent condition
angles <- phases_consistent[, 1]

# Circular mean direction
mu <- circ_mean(angles)
cat("Mean direction:", round(mu, 3), "rad\n")

# Mean resultant length (concentration)
r <- circ_r(angles)
cat("Resultant length:", round(r, 3), "\n")

# Rayleigh test for non-uniformity
rayleigh <- circ_rayleigh(angles)
cat("Rayleigh p-value:", format(rayleigh$pval, digits = 3), "\n")

# V-test against expected direction of 0
vtest <- circ_vtest(angles, dir = 0)
cat("V-test p-value:", format(vtest$pval, digits = 3), "\n")
```

### Visualizing Phase Distribution

```{r phase-histogram, fig.cap="Phase distribution showing clustering around 0"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_phase_hist(angles, nbins = 16)
}
```

## Cluster Detection on Time-Frequency Maps

Cluster-based permutation testing is the standard approach for multiple
comparison correction in time-frequency analyses. It works by:

1. Thresholding the statistical map to identify contiguous clusters
2. Computing a cluster statistic (sum of values)
3. Comparing against a null distribution from permutations

### Simulating Time-Frequency Data

```{r simulate-tf}
set.seed(789)

# Simulate a time-frequency map with an embedded effect cluster
n_freq <- 30
n_time <- 50
n_subjects <- 15

# Baseline condition: random noise
baseline <- array(
  rnorm(n_subjects * n_freq * n_time, mean = 0, sd = 1),
  dim = c(n_subjects, n_freq, n_time)
)

# Effect condition: noise + cluster of increased activity
effect <- array(
  rnorm(n_subjects * n_freq * n_time, mean = 0, sd = 1),
  dim = c(n_subjects, n_freq, n_time)
)

# Add effect cluster (theta band: rows 5-10, time 15-25)
effect[, 5:10, 15:25] <- effect[, 5:10, 15:25] + 1.5

cat("Data dimensions:", dim(effect), "(subjects x freq x time)\n")
```

### Computing T-scores

```{r compute-tscores}
# Paired t-test across subjects (dimension 1)
tres <- paired_tscore(effect, baseline, dim = 1)
t_matrix <- matrix(tres$t, nrow = n_freq, ncol = n_time)

cat("T-score range:", round(range(t_matrix, na.rm = TRUE), 2), "\n")
cat("Max t-score location: freq", which.max(apply(t_matrix, 1, max)),
    ", time", which.max(apply(t_matrix, 2, max)), "\n")
```

### Detecting Clusters

```{r detect-clusters}
# Define frequency and time axes
freq_axis <- seq(2, 30, length.out = n_freq)
time_axis <- seq(-0.5, 1.5, length.out = n_time)

# Detect clusters using extract method
clusters <- detect_clusters(
  data = t_matrix,
  threshold = 2.0,  # t-threshold (roughly p < 0.05 uncorrected)
  method = "extract",
  smooth = 1,
  freq_axis = freq_axis,
  time_axis = time_axis
)

cat("Number of clusters found:", length(clusters), "\n")

# Examine clusters
if (length(clusters) > 0) {
  for (i in seq_along(clusters)) {
    cl <- clusters[[i]]
    cat(sprintf("\nCluster %d:\n", i))
    cat(sprintf("  Frequency: %.1f - %.1f Hz\n", min(cl$freqs), max(cl$freqs)))
    cat(sprintf("  Time: %.2f - %.2f s\n", min(cl$times), max(cl$times)))
    cat(sprintf("  Size: %d pixels\n", length(cl$Zscores)))
    cat(sprintf("  Sum t: %.2f\n", cl$sumZscore))
    cat(sprintf("  Peak t: %.2f\n", cl$peakZscore))
  }
}
```

### Visualizing Clusters

```{r plot-clusters, fig.cap="Time-frequency map with detected clusters outlined"}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_cluster_heatmap(
    data = t_matrix,
    clusters = clusters,
    freq_axis = freq_axis,
    time_axis = time_axis,
    zlim = c(-4, 4)
  )
}
```

### Watershed vs Extract Methods

Two clustering methods are available:

| Method | Description | Best for |
|--------|-------------|----------|
| `extract` | 8-connectivity connected components | Sharply defined clusters |
| `watershed` | Watershed segmentation | Smooth, overlapping regions |

```{r watershed-comparison, eval=FALSE}
# Watershed method (requires smoothing for best results)
clusters_ws <- detect_clusters(
  data = t_matrix,
  threshold = 2.0,
  method = "watershed",
  smooth = c(2, 2),  # smooth in both dimensions
  freq_axis = freq_axis,
  time_axis = time_axis
)
```

## Multiple Comparison Correction

### FDR Correction

For element-wise testing without clustering, use FDR (False Discovery Rate):

```{r fdr-correction}
# Convert t-scores to p-values
df <- n_subjects - 1
p_matrix <- 2 * pt(-abs(t_matrix), df = df)

# Apply Benjamini-Hochberg FDR correction
fdr_result <- fdr_bh(as.vector(p_matrix), q = 0.05)

cat("Total tests:", length(p_matrix), "\n")
cat("Significant after FDR:", sum(fdr_result$h), "\n")
cat("Critical p-value:", format(fdr_result$crit_p, digits = 3), "\n")
```

### Non-parametric P-values

For permutation-based testing, use `nonparam_pval()`:

```{r nonparam-pval}
set.seed(111)

# Observed statistics (can be a vector)
observed_stats <- c(15.5, 12.0, 8.5)

# Null distribution from permutations
null_distribution <- rnorm(1000, mean = 10, sd = 3)

# Compute p-values
result <- nonparam_pval(observed_stats, null_distribution)
cat("Observed values:", observed_stats, "\n")
cat("P-values:", round(result$p, 4), "\n")
cat("Significant (alpha=0.05):", result$h, "\n")
```

## U-Score Matrix (Non-parametric Testing)

For comparing observed values against a permutation distribution across a matrix:

```{r uscore-example}
set.seed(222)

# Observed values (2D matrix)
observed <- matrix(rnorm(20, mean = 0.5), nrow = 4, ncol = 5)

# Reference distribution (100 permutations)
reference <- array(rnorm(100 * 4 * 5), dim = c(100, 4, 5))

# Compute U-scores (position in null distribution)
u_result <- u_score_matrix(observed, reference, alpha = 0.05)

cat("Significant elements:", sum(u_result$sgnf != 0), "of", length(observed), "\n")
```

## Complete Analysis Workflow

Here's a complete workflow combining PPC analysis with cluster detection:

```{r complete-workflow}
set.seed(333)

# 1. Simulate PPC data: correct vs incorrect trials
n_subjects <- 12
n_freq <- 20
n_time <- 30

# Correct trials: higher theta PPC around response time
ppc_correct <- array(
  runif(n_subjects * n_freq * n_time, 0, 0.3),
  dim = c(n_subjects, n_freq, n_time)
)
# Add theta effect (freqs 3-6 = theta, times 10-20 = peri-response)
ppc_correct[, 3:6, 10:20] <- ppc_correct[, 3:6, 10:20] + 0.25

# Incorrect trials: baseline PPC
ppc_incorrect <- array(
  runif(n_subjects * n_freq * n_time, 0, 0.3),
  dim = c(n_subjects, n_freq, n_time)
)

# 2. Statistical comparison
diff_t <- paired_tscore(ppc_correct, ppc_incorrect, dim = 1)$t
diff_matrix <- matrix(diff_t, nrow = n_freq, ncol = n_time)

# 3. Cluster detection
freq_ax <- seq(2, 30, length.out = n_freq)
time_ax <- seq(-0.2, 0.8, length.out = n_time)

sig_clusters <- detect_clusters(
  data = diff_matrix,
  threshold = 2.0,
  method = "extract",
  smooth = 1.5,
  freq_axis = freq_ax,
  time_axis = time_ax
)

# 4. Report
cat("=== PPC: Correct vs Incorrect Trials ===\n")
cat("Subjects:", n_subjects, "\n")
cat("Clusters found:", length(sig_clusters), "\n\n")

for (i in seq_along(sig_clusters)) {
  cl <- sig_clusters[[i]]
  cat(sprintf("Cluster %d:\n", i))
  cat(sprintf("  Frequency: %.1f - %.1f Hz\n", min(cl$freqs), max(cl$freqs)))
  cat(sprintf("  Time: %.2f - %.2f s\n", min(cl$times), max(cl$times)))
  cat(sprintf("  Center: %.1f Hz, %.2f s\n", cl$CoMfreq, cl$CoMtime))
  cat(sprintf("  Cluster t-sum: %.1f\n\n", cl$sumZscore))
}
```

## Loading Real Data

For actual experiments, use the I/O utilities:

```{r io-example, eval=FALSE}
# Load MATLAB .mat file (requires R.matlab package)
mat_data <- read_mat("path/to/ppc_data.mat")

# Access variables
ppc_matrix <- mat_data$ppc
freq_axis <- mat_data$freqs
time_axis <- mat_data$times

# Load HDF5 file (requires hdf5r package)
h5_file <- open_hdf5("path/to/data.h5")
ppc_matrix <- h5_file[["ppc"]][]
h5_file$close_all()
```

## Summary

| Function | Purpose |
|----------|---------|
| `pairwise_phase_consistency()` | Bias-free phase consistency measure |
| `circ_mean()`, `circ_r()` | Circular mean and resultant length |
| `circ_rayleigh()`, `circ_vtest()` | Significance tests for phase clustering |
| `narrowband_hilbert()` | Extract phase at specific frequency |
| `paired_tscore()` | Paired t-test across subjects |
| `detect_clusters()` | Find significant clusters in time-frequency maps |
| `fdr_bh()` | Benjamini-Hochberg FDR correction |
| `u_score_matrix()` | Non-parametric significance testing |
| `plot_phase_hist()` | Visualize phase distributions |
| `plot_cluster_heatmap()` | Visualize time-frequency clusters |

## Session Info

```{r session-info}
sessionInfo()
```
