---
title: "Workflow Wrappers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow Wrappers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
has_imager <- requireNamespace("imager", quietly = TRUE)
```

## Introduction

The **bosc** package provides low-level functions for oscillation analysis, but
also includes high-level workflow wrappers that simplify common analysis patterns.
These wrappers are designed to:
- Accept configuration lists for reproducible, programmatic analyses
- Combine multiple steps (e.g., compute score + surrogates + significance) in one call
- Provide tidy output formats for easy integration with tidyverse workflows

For the underlying algorithms, see `vignette("oscillation-score")` and
`vignette("ppc-clustering")`.

```{r load}
library(bosc)
```

## Quick Start: One-Call Significance Testing

The most common workflow is computing an oscillation score and testing its
significance against surrogates. `oscillation_score_z()` does this in one call:

```{r quick-start}
# Generate a signal with 10 Hz periodicity
set.seed(42)
fs <- 1000
sig <- numeric(fs * 2)
spike_times <- seq(50, 1950, by = 100) + round(rnorm(20, sd = 5))
spike_times <- spike_times[spike_times > 0 & spike_times <= length(sig)]
sig[spike_times] <- 1

# One-call analysis: compute oscore, surrogates, and z-score
result <- oscillation_score_z(
  signal = sig,
  fs = fs,
  flim = c(5, 20),
  nrep = 200,
  alpha = 0.05,
  taper = "hanning"
)

cat("Oscillation score:", round(result$oscore, 3), "\n")
cat("Peak frequency:", round(result$fosc, 2), "Hz\n")
cat("Z-score:", round(result$z, 2), "\n")
cat("P-value:", format(result$pval, digits = 3), "\n")
cat("Significant:", result$significant, "\n")
```

## Configuration-List Workflows

When running batch analyses across subjects or conditions, it's convenient to
define parameters once and reuse them. The `_config` wrappers accept named lists:

### Basic Config-Driven Analysis

```{r config-basic}
# Define analysis parameters
config <- list(
  fs = 1000,
  flim = c(1, 50),
  quantlim = c(0.05, 0.95),
  smoothach = TRUE,
  smoothwind = 0.002,
  peakwind = 0.008,
  taper = "hanning",
  warnings = "off"
)

# Apply to signal
os <- oscillation_score_config(config, sig)

cat("Oscillation score:", round(os$oscore, 3), "\n")
cat("Peak frequency:", round(os$fosc, 2), "Hz\n")
```

### Surrogate Config Workflow

```{r config-surrogates}
# Extend config for surrogate analysis
config_sur <- modifyList(config, list(
  nrep = 200,
  fpeak = os$fosc,
  keep_trend = FALSE
))

sur <- oscillation_score_surrogates_config(config_sur, sig)

# Compute statistics from surrogate distribution
z <- (log(os$oscore) - mean(log(sur$oscore_rp), na.rm = TRUE)) /
     sd(log(sur$oscore_rp), na.rm = TRUE)

cat("Surrogate mean:", round(mean(sur$oscore_rp), 3), "\n")
cat("Surrogate SD:", round(sd(sur$oscore_rp), 3), "\n")
cat("Log-Z score:", round(z, 2), "\n")
```

### Batch Analysis Example

```{r batch-analysis}
# Simulate multiple subjects with varying oscillation strength
set.seed(123)
n_subjects <- 5

# Create subjects with different oscillation strengths
subjects <- lapply(1:n_subjects, function(i) {
  sig <- numeric(2000)
  # Vary jitter: stronger oscillation = less jitter
  jitter_sd <- 5 + (i - 1) * 3
  spikes <- seq(50, 1950, by = 100) + round(rnorm(20, sd = jitter_sd))
  spikes <- spikes[spikes > 0 & spikes <= 2000]
  sig[spikes] <- 1
  sig
})

# Apply same config to all subjects
results <- lapply(subjects, function(s) {
  oscillation_score_z(s, fs = 1000, flim = c(5, 20), nrep = 100, tidy = TRUE)
})

# Combine results
batch_results <- do.call(rbind, results)
batch_results$subject <- 1:n_subjects
print(batch_results[, c("subject", "oscore", "fosc", "z", "significant")])
```

## Phase at Event Times

A common analysis extracts the instantaneous phase of an oscillation at specific
event times (e.g., button presses, stimulus onsets). `phase_at_events()` handles
the full pipeline:

1. Convert event times to a continuous trace

2. Narrowband filter at the frequency of interest
3. Extract Hilbert phase at each event

```{r phase-events}
# Event times in seconds (e.g., button presses)
events <- c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85, 0.95)

# Extract phase at 10 Hz
phases <- phase_at_events(
  events = events,
  dt = 0.001,      # 1 ms time resolution
  fosc = 10,       # Center frequency
  bandwidth = 1    # +/- 1 Hz band
)

cat("Phases at events (radians):\n")
print(round(phases, 2))

# Check phase consistency
cat("\nMean resultant length:", round(circ_r(phases), 3), "\n")
ray <- circ_rayleigh(phases)
cat("Rayleigh p-value:", format(ray$pval, digits = 3), "\n")
```

### Leave-One-Out Phase Estimation

For unbiased phase estimation (avoiding circular analysis), use leave-one-out:

```{r loo-phases, eval=FALSE}
# Each event's phase is computed from a trace excluding that event
phases_loo <- phase_at_events(
  events = events,
  dt = 0.001,
  fosc = 10,
  bandwidth = 1,
  leave_one_out = TRUE
)
```

## Tidy Output Helpers

The package includes helper functions to convert results to tidy data.frames,
making it easy to combine results across analyses or use with ggplot2/dplyr.
### oscore_tidy: Oscillation Score Summaries

```{r tidy-oscore}
# Single result
oscore_tidy(os)

# Can also handle oscillation_score_z output
z_result <- oscillation_score_z(sig, fs = 1000, flim = c(5, 20), nrep = 50)
oscore_tidy(z_result)
```

### oscore_spectrum: Extract Spectrum Data

```{r tidy-spectrum}
# Get spectrum as a data.frame for plotting
spec_df <- oscore_spectrum(os)
head(spec_df)

# Plot with ggplot2
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  ggplot(spec_df, aes(x = freq, y = power)) +
    geom_line() +
    geom_vline(xintercept = os$fosc, linetype = "dashed", color = "red") +
    labs(x = "Frequency (Hz)", y = "Power", title = "Oscillation Spectrum") +
    theme_minimal()
}
```

### clusters_tidy: Cluster Summaries

After detecting clusters in time-frequency maps, convert to tidy format:

```{r tidy-clusters, eval=has_imager}
# Create example time-frequency data with clusters
set.seed(456)
n_freq <- 20
n_time <- 30
t_matrix <- matrix(rnorm(n_freq * n_time), nrow = n_freq)
# Add cluster
t_matrix[5:8, 10:15] <- t_matrix[5:8, 10:15] + 3

# Detect clusters
freq_axis <- seq(2, 30, length.out = n_freq)
time_axis <- seq(-0.2, 0.8, length.out = n_time)

clusters <- detect_clusters(
  data = t_matrix,
  threshold = 2.0,
  method = "extract",
  smooth = 1,
  freq_axis = freq_axis,
  time_axis = time_axis
)

# Convert to tidy summary
cluster_summary <- clusters_tidy(clusters)
print(cluster_summary[, c("cluster", "n_pixels", "freq_min", "freq_max",
                          "time_min", "time_max", "sumZscore")])
```

### clusters_pixels: Per-Pixel Data

For detailed analysis or custom plotting, get all cluster pixels:

```{r tidy-pixels, eval=has_imager}
# Get all pixels from detected clusters
pixel_df <- clusters_pixels(clusters)
head(pixel_df)

# Example: plot cluster extent
if (requireNamespace("ggplot2", quietly = TRUE) && nrow(pixel_df) > 0) {
  ggplot(pixel_df, aes(x = time, y = freq, fill = Zscore)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    labs(x = "Time (s)", y = "Frequency (Hz)", fill = "Z-score",
         title = "Cluster Pixels") +
    theme_minimal()
}
```

## Complete Analysis Pipeline

Here's a complete example combining the workflow wrappers:

```{r complete-pipeline}
set.seed(789)

# 1. Generate behavioral data with oscillatory structure
fs <- 1000
duration <- 3
n_samples <- fs * duration

# Response times with 8 Hz periodicity
base_times <- seq(0.1, 2.9, by = 0.125)  # 8 Hz
event_times <- base_times + rnorm(length(base_times), sd = 0.01)
event_times <- sort(event_times[event_times > 0 & event_times < duration])

# Convert to spike train
sig <- numeric(n_samples)
sig[round(event_times * fs)] <- 1

# 2. Compute oscillation score with significance
config <- list(
  fs = fs,
  flim = c(4, 15),
  nrep = 200,
  alpha = 0.05,
  taper = "hanning"
)

z_result <- oscillation_score_z(
  signal = sig,
  fs = config$fs,
  flim = config$flim,
  nrep = config$nrep,
  alpha = config$alpha,
  taper = config$taper
)

# 3. Extract phases at response times
phases <- phase_at_events(
  events = event_times,
  dt = 1/fs,
  fosc = z_result$fosc,
  bandwidth = 1
)

# 4. Test phase consistency
ray <- circ_rayleigh(phases)

# 5. Report results
cat("=== Oscillation Analysis Results ===\n\n")

cat("Oscillation Score Analysis:\n")
cat(sprintf("  O-score: %.3f\n", z_result$oscore))
cat(sprintf("  Peak frequency: %.2f Hz\n", z_result$fosc))
cat(sprintf("  Z-score: %.2f\n", z_result$z))
cat(sprintf("  P-value: %.4f\n", z_result$pval))
cat(sprintf("  Significant: %s\n\n", z_result$significant))

cat("Phase Consistency:\n")
cat(sprintf("  Mean phase: %.2f rad\n", circ_mean(phases)))
cat(sprintf("  Resultant length: %.3f\n", circ_r(phases)))
cat(sprintf("  Rayleigh p-value: %.4f\n", ray$pval))
```

## Summary of Workflow Functions

| Function | Purpose | Key Benefit |
|----------|---------|-------------|
| `oscillation_score_z()` | O-score + surrogates + z-score | One-call significance testing |
| `oscillation_score_config()` | Config-driven O-score | Batch processing, reproducibility |
| `oscillation_score_surrogates_config()` | Config-driven surrogates | Batch processing |
| `phase_at_events()` | Extract phases at event times | Full pipeline in one call |
| `oscore_tidy()` | Convert results to data.frame | Tidyverse integration |
| `oscore_spectrum()` | Extract spectrum as data.frame | Easy plotting |
| `clusters_tidy()` | Summarize clusters | Quick cluster statistics |
| `clusters_pixels()` | Per-pixel cluster data | Custom analyses/plotting |

## Session Info

```{r session-info}
sessionInfo()
```
